A Mechanised Library for Generic (Co)Programming

### 2024-02-27 paper/paper.tex
Stuff

\documentclass[a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}
\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \addbibresource{references.bib}
\usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{mathpartir}
\usepackage{bussproofs}
\usepackage{xparse}
\usepackage[usenames, dvipsnames]{xcolor}
\usepackage{lipsum}
\usepackage{xargs}
\usepackage{hyperref}
\usepackage{tikz-cd}
\usepackage{todonotes}
\usepackage{url}
\usepackage{xspace}
\usepackage{rotating}
\usepackage{quiver}
\usepackage{minted}
\usepackage{newunicodechar}
\usepackage{microtype}
% \addbibresource{references.bib}
\usepackage{array}   % for \newcolumntype macro
\newcolumntype{C}{>{$}c<{$}} % math-mode version of "l" column type
\newcolumntype{L}{>{$}l<{$}} % math-mode version of "l" column type

@@

\author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Joan R. Public\footnote{Optional footnote, e.g. to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}

\authorrunning{J. Open Access and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

@@

\acknowledgements{I want to thank \dots}%optional

@@

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \mintinline{Coq}{Valid} that
determines whether a position is valid in a shape.
\begin{minted}{Coq}
Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
Record App (X : Type) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{minted}
This approach solves the the problem of the unsafe casts, since the code
extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml singleton
type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{minted}{Coq}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

This functor is built as the sum of two functors: $X \mapsto \mathbb{N}$ and $X \mapsto
X \times X$. Consider first $X \mapsto \mathbb{N}$. This is a constant functor
that maps $X$ to the type $\mathbb{N}$. More generally, a functor that, given a
type $A$, maps everything to it, $X \mapsto A$, only needs one shape ($A$, L9
of Fig~\ref{fig:poly}), and zero positions (In Coq, \mintinline{Coq}{False},
L10 of Fig~\ref{fig:poly}). The reason why there are zero valid positions in
a constant container extension is because, given something of type 
\mintinline{Coq}{App (Const (Ks A) Kp) X}, it should be impossible to get
something of type \mintinline{Coq}{X}. But from the container shape, we will
always be able to get the element of type $A$\footnote{%
  Our decision to use \mintinline{Coq}{False} is arbitrary. In general, we
  could have chosen any time, with a validity predicate that always returns
  \mintinline{Coq}{false}. It would make no difference for code extraction.
}. This functor is, therefore \mintinline{Coq}{Const nat}.

Consider now $X \mapsto X \times X$. This is the product of two
identity functors $X \mapsto X$. The identity functor only has one shape, and
we can use the unit type to represent such shapes. From this shape we should be
able to get an element of type $X$, and we can capture this by using the unit
type for positions. To build an element of the container extension type, we
will need to provide a function of type \mintinline{Coq}{unit -> X}, which is
isomorphic to \mintinline{Coq}{X}.

The shape of the product of two container extensions is the product of their
shapes. The product of such container extensions should allow us to get either
the left, or the right element in the product. Therefore, the positions of this
container extension is the sum of the positions of the two container
extensions. What is a valid position in a given shape? If the position is
\mintinline{Coq}{inl p} (or \mintinline{Coq}{inr p}), then it is only valid if
\mintinline{Coq}{p} is valid in the first (resp. second) component of the
shape. The product of our two container extensions is
\mintinline{Coq}{Prod Id Id}.

Finally, the shape the sum of two container extensions is the sum of their
shapes. The positions are, like for products, the sum of their shapes. But to
check the validity of shapes in positions, their constructor must match, e.g.
shape \mintinline{Coq}{inl s} and shape \mintinline{Coq}{inl p}, with
\mintinline{Coq}{valid s p}.

Our final container extension is the following: 
\mintinline{Coq}{Sum (Const nat) (Prod Id Id)}.
The shape of this functor is: 
\mintinline{Coq}{nat + unit * unit}. The positions of this functor
are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
to \mintinline{Coq}{unit + unit}, which will only be valid in the RHS.

The Coq value \mintinline{Coq}{inl n} of type
\mintinline{Coq}{nat + bool * bool} would look as follows as a container
extension 
\mintinline{Coq}{App (Sum (Const nat) (Prod Id Id)) bool}:
\begin{minted}{Coq}
{| shape := inl n; cont := fun e => uninhabited e|}
\end{minted}
We omit the definition of \mintinline{Coq}{uninhabited}, which exploits the
fact that \mintinline{Coq}{e} for \mintinline{Coq}{inl n} implies
\mintinline{Coq}{False} to build an element of type \mintinline{Coq}{bool}%
\footnote{Note that, since \mintinline{Coq}{e} is uninhabited we could return
any arbitrary value instead of \mintinline{Coq}{uninhabited e}.}.

Similarly, the Coq value \mintinline{Coq}{inr (true, false)} as a container
extension is as follows:
\begin{minted}{Coq}
  {| shape := inr (tt,tt);
     cont := fun e : {p | valid (inr (tt, tt)) p} =>
        match proj1_sig e with
        | inl f => match f with end
        | inr (inl tt) => fun _ => true
        | inr (inr tt) => fun _ => false
        end
  |}
\end{minted}
Here, the position \mintinline{Coq}{inl f} is impossible, as
\mintinline{Coq}{f} would need to be a proof of \mintinline{Coq}{False}.
Of the two remaining positions, the left one will return 
\mintinline{Coq}{true}, and the right position \mintinline{Coq}{false}.
\begin{figure}
\begin{minipage}{.52\textwidth}
  \begin{minted}[linenos, fontsize=\small]{Coq}
Context `(C1 : Cont S1 P1) 
        `(C2 : Cont S2 P2).

@@

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \mintinline{Coq}{App F A -> B} (or \mintinline{Coq}{B -> App F A}). Since 
these types now come with their own equivalence, any function that manipulates
them needs to be \emph{respectful}. I.e.\ given 
\mintinline{Coq}{R : X -> X -> Prop} and \mintinline{Coq}{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{minted}{Coq}
forall (x y : X), R x y -> R' (f x) (f y)
\end{minted}

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by coupling every type with an
associated equivalence relation, and every morphism with a proof that it is
\emph{respectful} (i.e. \emph{proper} morphisms with respect to the function
\emph{respectfulness} relation). We use \mintinline{Coq}{=e} to denote the
equivalence relation associated with a type. Every Coq type is associated by
default with the standard Coq propositional equality, unless a different
equivalence is specified. We use the type-class mechanism to help Coq's code
extraction mechanism to remove any occurence of custom equivalence relations
in the extracted OCaml code.
\begin{minted}{Coq}
Reserved Notation "f =e g" (at level 70, no associativity).
Class equiv A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{minted}
Given types \mintinline{Coq}{A} and  \mintinline{Coq}{B}, with their respective
equivalence relations \mintinline{Coq}{eA : A -> A -> Prop} and
\mintinline{Coq}{eB : B -> B -> Prop}, the we define the type
\mintinline{Coq}{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \mintinline{Coq}{R} to \mintinline{Coq}{R'}.
\begin{minted}{Coq}
Structure morph A {eA : equiv A} B {eB : equiv B} :=
  MkMorph
    { app :> A -> B;
      app_eq : forall x y, x =e y -> app x =e app y
    }.
Notation "A ~> B" := (@morph A _ B _).
\end{minted}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \mintinline{Coq}{Prop} in the code, so objects of type
\mintinline{Coq}{A ~> B} will be extracted to the OCaml equivalent to
\mintinline{Coq}{A -> B}.

\subsubsection{Polynomial Types}
We define a number of equivalences for polynomial types. These equivalences are
be used when building proper morphisms that manipulate their respective types.
\begin{minted}{Coq}
Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
Instance prop_eq : equiv Prop.
Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
\end{minted}

Most of the definitions that involve functions and polynomial types are
straightworward.  Identity and composition are defined as 
\mintinline{Coq}{fun x => x} and \mintinline{Coq}{fun f g x => f (g x)} 
respectively, and the proofs that they are proper morphisms is straightforward,
and automatically discharged by Coq. 
Products are built using function \mintinline{Coq}{fun f g x => (f x, g x)},
with the projections being the standard Coq  \mintinline{Coq}{fst} and
\mintinline{Coq}{snd} functions. Similarly, sum injections are encoded using
Coq's \mintinline{Coq}{inl} and \mintinline{Coq}{inr} constructors, and
pattern matching on them uses the function 
\mintinline{Coq}{fun f g x => match x with | inl y => f y | inr y => g y end}.
The proofs that these morphisms are proper are straightforward.  Finally, we
also provide functions for currying/uncurrying, and partially applying a proper
morphism. We avoid using definitions, or unfold them, to help Coq's code
extraction mechanism to inline as many of these small definitions as possible.

@@

Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{minted}{Coq}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape ~> Pos ~> bool }.
Inductive AppR `{e : equiv X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{minted}
Note the use of \mintinline{Coq}{=e} instead of Coq's standard equality.  For
positions, however, we chose to use Coq's propositional equality, since these
would leads to simpler code. We explain why in Section~\ref{sec:cata}, and the
mechanisation of initial algebras of container extensions. The definitions in
Figure~\ref{fig:poly} are redefined to account for these changes in a
straightforward way.

We prove the isomorphisms of polynomial types and the equivalent (redefined)
container extensions. As an example, we discuss (informally) the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \mintinline{Coq}{App F X} is isomorphic to \mintinline{Coq}{A}, and
\mintinline{Coq}{App G X} is isomorphic to \mintinline{Coq}{B}. Then we can
show that \mintinline{Coq}{App (Prod F G) X} is isomorphic to
\mintinline{Coq}{A * B}.  If we have an element of type 
\mintinline{Coq}{App (Prod F G) X},
using the \mintinline{Coq}{inl} position, we can obtain
\mintinline{Coq}{App F X}.  Similarly, using \mintinline{Coq}{inr}, we can
obtain \mintinline{Coq}{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \mintinline{Coq}{App F X} and \mintinline{Coq}{App G X} to
obtain \mintinline{Coq}{A * B}. Similarly if we have \mintinline{Coq}{A * B},
we can first use the isomorphisms of 
\mintinline{Coq}{A} and
\mintinline{Coq}{B} to obtain
\mintinline{Coq}{App F X * App G X}, and then construct the necessary container
extension. Given \mintinline{Coq}{p_inl}/\mintinline{Coq}{p_inr} that act as
\mintinline{Coq}{inl/inr} but on valid product positions, the relevant
functions that define the isomorphism are:
\begin{minted}{Coq}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (e_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (e_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := fun e => match val e with 
                    | inl p => fun H => cont (fst x) (ValidPos p H)
                    | inr p => fun H => cont (snd x) (ValidPos p H)
                    end (IsValid e)
|}.
\end{minted}
Proving that these functions indeed form an isomorphism is straightforward,
using the fact that Product containers only have two valid positions.

@@

Algebras on container extensions are defined as expected. The carrier of the
algebra is \mintinline{Coq}{A}, and the operations of the algebra are morphisms
of the following type:
\begin{minted}{Coq}
Notation Alg F A := (App F A ~> A).
\end{minted}
Given a container extension \mintinline{Coq}{F}, we define \emph{two}
(isomorphic) initial algebras. We will show the need for the second initial
algebra in the following section. Our fist initial algebra is defined as the
straightforward least fixpoint of the container extension \mintinline{Coq}{F}:
\begin{minted}{Coq}
Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
\end{minted}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words,
we would need to provide an equivalence for \mintinline{Coq}{LFix}, but the
only way to do this is to define such equivalence simultaneously to the
definition for  \mintinline{Coq}{LFix}.  We could have overcame this
circularity by defining  \mintinline{Coq}{LFix} independently of container
extensions, and define later the necessary
\mintinline{Coq}{in}/\mintinline{Coq}{out} functions to convert between
\mintinline{Coq}{LFix} and \mintinline{Coq}{App}. However, this extra layer
interferes with code extraction, in that it requires destructing/constructing
subsequent \mintinline{Coq}{LFix} and  \mintinline{Coq}{App}, which Coq is not
always able to inline.

>>>>

\documentclass[anonymous, a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}
\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \addbibresource{references.bib}
\usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{mathpartir}
\usepackage{bussproofs}
\usepackage{xparse}
\usepackage[usenames, dvipsnames]{xcolor}
\usepackage{lipsum}
\usepackage{xargs}
\usepackage{hyperref}
\usepackage{tikz-cd}
\usepackage{todonotes}
\usepackage{url}
\usepackage{xspace}
\usepackage{rotating}
\usepackage{quiver}
\usepackage{minted}
\usepackage{newunicodechar}
\usepackage{microtype}
% \addbibresource{references.bib}
\usepackage{array}   % for \newcolumntype macro
\newcolumntype{C}{>{$}c<{$}} % math-mode version of "l" column type
\newcolumntype{L}{>{$}l<{$}} % math-mode version of "l" column type

@@

\author{David Castro-Perez}{School of Computing, University of Kent}{d-castro-perez@kent.ac.uk}{}{}
\author{Marco Paviotti}{School of Computing, University of Kent}{m.paviotti@kent.ac.uk}{}{}
\author{Michael Vollmer}{School of Computing, University of Kent}{m.vollmer@kent.ac.uk}{}{}

%\author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

%\author{Joan R. Public\footnote{Optional footnote, e.g. to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}

\authorrunning{D. Castro-Perez, M. Paviotti, and M. Vollmer} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{D. Castro-Perez, M. Paviotti, and M. Vollmer} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\keywords{hylomorphisms, Coq, code extraction, verification} %TODO mandatory; please add comma-separated list of keywords

@@

%\acknowledgements{I want to thank \dots}%optional

@@

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \coq{Valid} that
determines whether a position is valid in a shape.
\begin{minted}{Coq}
Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
Record App (X : Type) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{minted}
This approach solves the the problem of the unsafe casts, since the code
extracted for \coq{{p | Valid shape pos}} will be an OCaml singleton
type defined as the OCaml equivalent to \coq{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \coq{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{minted}{Coq}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

This functor is built as the sum of two functors: $X \mapsto \mathbb{N}$ and $X \mapsto
X \times X$. Consider first $X \mapsto \mathbb{N}$. This is a constant functor
that maps $X$ to the type $\mathbb{N}$. More generally, a functor that, given a
type $A$, maps everything to it, $X \mapsto A$, only needs one shape ($A$, L9
of Fig~\ref{fig:poly}), and zero positions (In Coq, \coq{False},
L10 of Fig~\ref{fig:poly}). The reason why there are zero valid positions in
a constant container extension is because, given something of type 
\coq{App (Const (Ks A) Kp) X}, it should be impossible to get
something of type \coq{X}. But from the container shape, we will
always be able to get the element of type $A$\footnote{%
  Our decision to use \coq{False} is arbitrary. In general, we
  could have chosen any time, with a validity predicate that always returns
  \coq{false}. It would make no difference for code extraction.
}. This functor is, therefore \coq{Const nat}.

Consider now $X \mapsto X \times X$. This is the product of two
identity functors $X \mapsto X$. The identity functor only has one shape, and
we can use the unit type to represent such shapes. From this shape we should be
able to get an element of type $X$, and we can capture this by using the unit
type for positions. To build an element of the container extension type, we
will need to provide a function of type \coq{unit -> X}, which is
isomorphic to \coq{X}.

The shape of the product of two container extensions is the product of their
shapes. The product of such container extensions should allow us to get either
the left, or the right element in the product. Therefore, the positions of this
container extension is the sum of the positions of the two container
extensions. What is a valid position in a given shape? If the position is
\coq{inl p} (or \coq{inr p}), then it is only valid if
\coq{p} is valid in the first (resp. second) component of the
shape. The product of our two container extensions is
\coq{Prod Id Id}.

Finally, the shape the sum of two container extensions is the sum of their
shapes. The positions are, like for products, the sum of their shapes. But to
check the validity of shapes in positions, their constructor must match, e.g.
shape \coq{inl s} and shape \coq{inl p}, with
\coq{valid s p}.

Our final container extension is the following: 
\coq{Sum (Const nat) (Prod Id Id)}.
The shape of this functor is: 
\coq{nat + unit * unit}. The positions of this functor
are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
to \coq{unit + unit}, which will only be valid in the RHS.

The Coq value \coq{inl n} of type
\coq{nat + bool * bool} would look as follows as a container
extension 
\coq{App (Sum (Const nat) (Prod Id Id)) bool}:
\begin{minted}{Coq}
{| shape := inl n; cont := fun e => uninhabited e|}
\end{minted}
We omit the definition of \coq{uninhabited}, which exploits the
fact that \coq{e} for \coq{inl n} implies
\coq{False} to build an element of type \coq{bool}%
\footnote{Note that, since \coq{e} is uninhabited we could return
any arbitrary value instead of \coq{uninhabited e}.}.

Similarly, the Coq value \coq{inr (true, false)} as a container
extension is as follows:
\begin{minted}{Coq}
  {| shape := inr (tt,tt);
     cont := fun e : {p | valid (inr (tt, tt)) p} =>
        match proj1_sig e with
        | inl f => match f with end
        | inr (inl tt) => fun _ => true
        | inr (inr tt) => fun _ => false
        end
  |}
\end{minted}
Here, the position \coq{inl f} is impossible, as
\coq{f} would need to be a proof of \coq{False}.
Of the two remaining positions, the left one will return 
\coq{true}, and the right position \coq{false}.
\begin{figure}
\begin{minipage}{.52\textwidth}
  \begin{minted}[linenos, fontsize=\small]{Coq}
Context `(C1 : Cont S1 P1) 
        `(C2 : Cont S2 P2).

@@

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \coq{App F A -> B} (or \coq{B -> App F A}). Since 
these types now come with their own equivalence, any function that manipulates
them needs to be \emph{respectful}. I.e.\ given 
\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{minted}{Coq}
forall (x y : X), R x y -> R' (f x) (f y)
\end{minted}

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of setoids,
i.e.\ types with na associated equivalence relation, and considering only
functions that respect the associated equivalences, i.e. \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation.  We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code.  We use \coq{=e} to denote
the equivalence relation of a setoid. By default, we associate every Coq type
with the standard propositional equality, unless a different equivalence is
specified (i.e.\ we allow overlapping instances, and Coq's propositional
equality takes the lowest priority). 
\begin{minted}{Coq}
Reserved Notation "f =e g" (at level 70, no associativity).
Class setoid A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{minted}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{minted}{Coq}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph
    { app :> A -> B;
      app_eq : forall x y, x =e y -> app x =e app y
    }.
Notation "A ~> B" := (@morph A _ B _).
\end{minted}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \coq{Prop} in the code, so objects of type
\coq{A ~> B} will be extracted to the OCaml equivalent to
\coq{A -> B}.

\subsubsection{Polynomial Types}
We define a number of equivalences for polynomial types. These equivalences are
be used when building proper morphisms that manipulate their respective types.
\begin{minted}{Coq}
Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
Instance prop_eq : setoid Prop.
Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
\end{minted}

Most of the definitions that involve functions and polynomial types are
straightworward.  Identity and composition are defined as 
\coq{fun x => x} and \coq{fun f g x => f (g x)} 
respectively, and the proofs that they are proper morphisms is straightforward,
and automatically discharged by Coq. 
Products are built using function \coq{fun f g x => (f x, g x)},
with the projections being the standard Coq  \coq{fst} and
\coq{snd} functions. Similarly, sum injections are encoded using
Coq's \coq{inl} and \coq{inr} constructors, and
pattern matching on them uses the function 
\coq{fun f g x => match x with | inl y => f y | inr y => g y end}.
The proofs that these morphisms are proper are straightforward.  Finally, we
also provide functions for currying/uncurrying, and partially applying a proper
morphism. We avoid using definitions, or unfold them, to help Coq's code
extraction mechanism to inline as many of these small definitions as possible.

@@

Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{minted}{Coq}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape ~> Pos ~> bool }.
Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{minted}
Note the use of \coq{=e} instead of Coq's standard equality.  For
positions, however, we chose to use Coq's propositional equality, since these
would leads to simpler code. We explain why in Section~\ref{sec:cata}, and the
mechanisation of initial algebras of container extensions. The definitions in
Figure~\ref{fig:poly} are redefined to account for these changes in a
straightforward way.

We prove the isomorphisms of polynomial types and the equivalent (redefined)
container extensions. As an example, we discuss (informally) the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl}/\coq{p_inr} that act as
\coq{inl/inr} but on valid product positions, the relevant
functions that define the isomorphism are:
\begin{minted}{Coq}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (e_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (e_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := fun e => match val e with 
                    | inl p => fun H => cont (fst x) (ValidPos p H)
                    | inr p => fun H => cont (snd x) (ValidPos p H)
                    end (IsValid e)
|}.
\end{minted}
Proving that these functions indeed form an isomorphism is straightforward,
using the fact that Product containers only have two valid positions.

@@

Algebras on container extensions are defined as expected. The carrier of the
algebra is \coq{A}, and the operations of the algebra are morphisms
of the following type:
\begin{minted}{Coq}
Notation Alg F A := (App F A ~> A).
\end{minted}
Given a container extension \coq{F}, we define \emph{two}
(isomorphic) initial algebras. We will show the need for the second initial
algebra in the following section. Our fist initial algebra is defined as the
straightforward least fixpoint of the container extension \coq{F}:
\begin{minted}{Coq}
Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
\end{minted}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words,
we would need to provide an equivalence for \coq{LFix}, but the
only way to do this is to define such equivalence simultaneously to the
definition for  \coq{LFix}.  We could have overcame this
circularity by defining  \coq{LFix} independently of container
extensions, and define later the necessary
\coq{in}/\coq{out} functions to convert between
\coq{LFix} and \coq{App}. However, this extra layer
interferes with code extraction, in that it requires destructing/constructing
subsequent \coq{LFix} and  \coq{App}, which Coq is not
always able to inline.
